A Plasma Torus Around a Young Low-Mass Star

### 2025-03-13 paper/ms.tex
update ms

{\it Radial Velocity}---We measured radial velocities of TIC~141146667
from each of our spectra using a pipeline that we developed specifically
for rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our absolute
velocity zero-points as well as viable orders for velocity measurement
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis pipeline reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

For TIC~141146667, we measured the radial velocities from each of our
HIRES spectra using regions near the K~I (7700\,\AA) resonance line and
three TiO bandheads (5160\,\AA, 5450\,\AA, and 5600\,\AA).  These
regions were selected because they provided the best matches between the
synthetic and observed spectra.  The resulting redshift measurements
over each order were averaged to produce the final measurement.  We used
the scatter of resulting velocity measurements between orders to assign
the RV uncertainty at each epoch.  The uncertainty-weighted mean
systemic velocity over all epochs on UT 2024-02-17 was
$\gamma$=0.6$\pm$1.5\,\kms.  The relative radial velocities about this
mean are given in Table~\ref{tab:rv}.

@@

{\it Age: No Obvious Association Membership}---Previous work
\cite{Bouma2024} has found that over 90\% of CPVs are associated with
known young moving groups based on their positions and kinematics.
TIC~141146667 is one of the exceptions.  We calculated the probability
of TIC~141146667 being part of any of the nearby known groups using
BANYAN\,$\Sigma$ v1.2 \cite{Gagne2018}.  That particular model
classifies it as a field star at $>$99.9\% confidence.  We also searched the
local vicinity of TIC~141146667 for neighbors with similar projected on-sky
velocities using \texttt{comove} \cite{Tofflemire2021}.  This yielded
no strong candidates for co-moving stars --- defined as those with
projected tangential velocities $\Delta v_{\rm T}$$<$5\,\kms ---
that share its isochronal youth.

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main-sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.
{\bf Figure~X} shows the location of the star in the color--absolute
magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
Upper-Centaurus-Lupus (UCL), IC~2602, and the Pleiades.
To make this diagram, {\bf TODO}we adopted the USco and UCL members
from X, the IC~2602 members from Y, and the Pleiades members from Z,
and applied the same reddening corrections as in \cite{Bouma2022}.
To measure the isochrone age of TIC~141146667, we then followed the 
empirical approach proposed by \cite{Gagne2020} and
implemented in \cite{Bouma2022}.
The idea is to estimate the star (or star cluster's) age by
interpolating between the observed isochrones of star clusters with
ages that have been otherwise calibrated, for instance using
the lithium depletion-boundary method.
While this circumvents known modelling issues in molecular opacities
and starspot coverage for young M dwarfs,
the ages returned by this procedure depend on the ages
assumed for each reference cluster.
In this case, the most important reference is UCL, since TIC~141146667
overlaps directly on its isochrone on the CAMD.
We adopted the 16~Myr age for UCL from \cite{PecautMamajek2016}, which
is consistent with ages determined by more recent sub-clustering
analyses \cite{Ratzenbock2023} for the largest sub-populations.
This in turns yields an isochronal age for TIC~141146667 of
$t_{\rm iso}$=$16^{+19}_{-6}$\,Myr.
Qualitatively, this range of uncertainty is consistent with no stars
in the $\approx$40\,Myr IC~2602 cluster with similar redness as
TIC~141146667 being equally bright.

{\it Age: Lithium}--- Frankly, it's not obviously there!  (but maybe
this is bc of the blending, and a pEW would show smth small)
%TODO TODO TODO

% We have K=10.47, d=57pc - > DM ~= 3.78 -> M_K ~= 6.67
Figure7 of Wood23 suggests from the Feiden Li depletion curves that no lithium would imply $t$>37Myr.
However Figure5 frmo the same work suggests maybe try binning up the spectrum, and then comparing against Bochanski+2007 for a lithium-free template.

Try differential comparison against Barnard's star...

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable;
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity; {\it
iii)} weak lithium absorption.

{\it Effective temperature, radius, mass, and spectral type}
For the stellar effective temperature and radius, we adopted the results
from the spectral energy distribution fits described by
\cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we used the broadband magnitudes from Gaia DR2, APASS,
2MASS, SDSS, and WISE $W1$ and $W2$.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are typically only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
against the photometric standards presented by \cite{Bochanski2007};
this qualitatively agrees with the effective temperature from the SED
fitting \cite{Pecaut2013}.  With the effective temperature, stellar
radius, and previously determined age, we then estimated the stellar
mass by interpolating against the PARSEC v1.2S isochrones
\cite{Chen2014}.  Specifically, we used the distance metric defined in
Equation~8 of \cite{Bouma2024} to select the model mass closest to the
observed temperature, radius, and age.  We derived the uncertainty in
the mass based by propagating the statistical uncertainties in each of
these parameters.

>>>>

{\it Radial Velocity}---We measured radial velocities of TIC~141146667
our HIRES spectra using a pipeline that we developed specifically
for rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our pipeline
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis code reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

For TIC~141146667, we measured the radial velocities using regions
near the K~I (7700\,\AA) resonance line and three TiO bandheads
(5160\,\AA, 5450\,\AA, and 5600\,\AA).  We
selected these regions because they provided the best matches between the
synthetic and observed spectra.  We then averaged the resulting redshift measurements
over each order to produce the final measurement.  We used
the scatter of resulting velocity measurements between orders to assign
the RV uncertainty at each epoch.  The uncertainty-weighted mean
systemic velocity over all epochs on UT 2024-02-17 was
$\gamma$=0.6$\pm$1.5\,\kms.  The relative radial velocities about this
mean are given in Table~\ref{tab:rv}.

@@

{\it Effective temperature, radius, mass, and spectral type}
We determined the stellar effective temperature and radius by adopting
the results from the spectral energy distribution fitting procedure
described by \cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we used the broadband magnitudes from Gaia DR2, APASS,
2MASS, SDSS, and WISE $W1$ and $W2$.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are typically only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
against the photometric standards presented by \cite{Bochanski2007};
this qualitatively agrees with the effective temperature from the SED
fitting \cite{Pecaut2013}.  With the effective temperature, stellar
radius, and previously determined age, we then estimated the stellar
mass by interpolating against the PARSEC v1.2S isochrones
\cite{Chen2014}.  Specifically, we used the distance metric defined in
Equation~8 of \cite{Bouma2024} to select the model mass closest to the
observed temperature, radius, and age.  We derived the uncertainty in
the mass based by propagating the statistical uncertainties in each of
these parameters.

{\it Age: No Obvious Association Membership}---Previous work
\cite{Bouma2024} has found that over 90\% of CPVs are associated with
known young moving groups based on their positions and kinematics.
TIC~141146667 is in the minority.  We calculated the probability
of TIC~141146667 being part of any of the nearby known groups using
BANYAN\,$\Sigma$ v1.2 \cite{Gagne2018}.  That particular model
classifies it as a field star at $>$99.9\% confidence.  We also searched the
local vicinity of TIC~141146667 for neighbors with similar projected on-sky
velocities using \texttt{comove} \cite{Tofflemire2021}.  This yielded
no strong candidates for co-moving stars with
projected tangential velocities $\Delta v_{\rm T}$$<$5\,\kms
that share its isochronal youth.

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main-sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.
{\bf Figure~X} shows the location of the star in the color--absolute
magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
Upper-Centaurus-Lupus (UCL), IC~2602, and the Pleiades.
To make this diagram, {\bf TODO}we adopted the USco and UCL members
from X, the IC~2602 members from Y, and the Pleiades members from Z,
and applied the same reddening corrections as in \cite{Bouma2022}.
To measure the isochrone age of TIC~141146667, we then followed the 
empirical approach proposed by \cite{Gagne2020} and
implemented in \cite{Bouma2022}.
The idea is to estimate the star (or star cluster's) age by
interpolating between the observed isochrones of star clusters with
ages that have been otherwise calibrated, for instance using
the lithium depletion-boundary method.
While this circumvents known modelling issues in molecular opacities
and starspot coverage for young M dwarfs,
the ages returned by this procedure depend on the ages
assumed for each reference cluster.
In this case, the most important reference is UCL, since TIC~141146667
overlaps directly on its isochrone on the CAMD.
We adopted the 16~Myr age for UCL from \cite{PecautMamajek2016}, which
is consistent with ages determined by more recent sub-clustering
analyses \cite{Ratzenbock2023} for the largest sub-populations.
This in turns yields an isochronal age for TIC~141146667 of
$t_{\rm iso}$=$16^{+19}_{-6}$\,Myr.
Qualitatively, this range of uncertainty is consistent with no stars
in the $\approx$40\,Myr IC~2602 cluster with similar redness as
TIC~141146667 being equally bright.
%FIXME: update discussion following consideration of anomalous CPV
% brightnesses and colors...

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been
studied for over sixty years
\cite{Hayashi1963,Bildsten1997,Burke2004}.  \cite{Wood2023} provided
an instructive recent overview: an abbreviated summary is that cooler
and younger M dwarfs can show the 6707.8\,\AA\ doublet in absorption,
$\gtrsim$10\%
below their ``continua''.  Unlike Sun-like stars, the continuum for M
dwarfs is poorly defined due to their molecular absorption.  We
attempted a lithium measurement by constructing a
wavelength-binned and Doppler-corrected TIC~141146667 spectrum,
assigning its uncertainties based on the measured scatter across the
five hour dataset.  We then compared this average spectrum against the
nearest matching template from \cite{Bochanski2007}.  The data show a
slight depression near the
expected lithium wavelength, potentially consistent with a $\Delta
\lambda$$\approx$3\,\AA\ line broadening.  This feature nominally
yields ${\rm EW}_{\rm Li}$=71$^{+18}_{13}$\,m\AA, where the statistical
uncertainties are evaluated using a bootstrap resampling technique
from the statistical uncertainties in the spectrum.  Unfortunately,
the systematic uncertainties associated with the continuum
normalization are large, comparable to the amplitude of the feature;
we therefore prefer to quote the result of this measurement as a
$2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.
% FIXME: this is perhaps a hokey way to do it.  Better to quantify the
% systematic uncertainty, and incorporate that into your quoted
% value...

Desite uncertainties in the details, what can be stated with
confidence is that lithium is not abundant in the spectrum of
TIC~141146667.  Figure~\ref{fig:liew_population} compares our upper
limit against the lithium equivalent width measurements and
temperatures reported by \cite{Jeffries2023} based on the Gaia-ESO
spectroscopic survey.  If the star were $\lesssim$20\,Myr old, given
its temperature we would expect to see lithium in abundance
($\gtrsim$400\,m\AA).  Since we do not, we conclude that $t_{\rm
Li,emp}$$\gtrsim$20\,Myr.  The \cite{Feiden2016} isochrones provide a
point for theoretical comparison, suggesting that since $M_{\rm
K}$$=$6.67\,mag, $t_{\rm Li,th}$$\gtrsim$35\,Myr for complete
depletion.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf2.pdf}
  \caption[]{todo... }
  \label{fig:liew_population}
\end{figure}

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity.  The
first item constitutes evidence for youth since all known CPVs are
$\lesssim$200\,Myr \cite{Bouma2024}.  Reference sequences for
$\approx$40-50\,Myr clusters show no stars of equal luminosity and the
same temperature, suggesting a photometric isochronal age upper limit
of 50\,Myr.  The weak lithium absorption suggests an age of at least
20\,Myr based on an empirical comparison using Gaia-ESO spectra, or at
least 35\,Myr based on the \cite{Feiden2016} isochrones.  These
considerations yield our adopted age range of 20-50\,Myr.
